Supplementary Information

### 2021-05-26 doc/manuscript/ForestGEO_treerings.Rmd
fixing up intro, table 1

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
In the following paragraphs, we outline major questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, commonly over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].

Tree diameter, DBH, scales predictably with numerous traits affecting tree physiology and growth [e.g., height, crown size and position, root mass, hydraulic architecture, @couvreur_water_2018; @enquist_global_2002; @niklas_plant_2004] and therefore is itself linked to growth [e.g., @muller-landau_testing_2006; @foster_predicting_2016] and its climate sensitivity [e.g., @bennett_larger_2015; @mcgregor_tree_2020].
Yet, there remain inconsistencies as to what is considered a typical growth pattern. 
Tree ring records commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increases with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open, as opposed to in the understory (Table `r hypotheses`).
Building upon observed ontogenetic patterns in RW, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions such as respiration [**REFS**] and reproduction [@thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Solid understanding of these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

Beyond the direct effects of climate, rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns that may not be universal in forest settings, as discussed above. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, even after correcting for ontogenetic growth trends, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019], fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all points throughout the history of a dynamically changing stand.
Stand dynamics strongly influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, affecting ecosystem level patterns of C allocation. 
Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020]. 
Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past couple centuries) will generally decline, whereas populations sampled from largely undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).

Here, we develop, and apply to ten forested sites spanning 52 degrees latitude, a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
This approach allows us to address the following broad questions concerning and associated hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) for a globally distributed tree-ring dataset:
(1) How does interannual climate variation shape annual tree growth?
(2) How does growth rate vary with DBH?
(3) How have size-corrected growth rates changed through time?

@@

hypotheses_table <- subset(hypotheses_table, select = -c(Q, H, category,notes) )#remove columns that shouldn't be included
ft <- flextable(hypotheses_table)
ft <- italic(ft, i=question_rows, j=1)
ft <- bold(ft, i=c(question_rows, hypothesis_rows))
ft <- line_spacing(ft, space = 1, part = "all")
ft <- padding(ft, padding.top = 2)
ft <- padding(ft, padding.bottom = 2)
ft <- padding(ft, j = 1, padding.left = 20)
ft <- padding(ft, i = hypothesis_rows, j = 1, padding.left = 10)
ft <- padding(ft, i = question_rows, j = 1, padding.left = 0)
ft <- autofit(ft)
ft <- footnote( ft, i = 1, j = 2,
                value = as_paragraph(c("SSC= species-site combinations; SSCV=species-site-variable combination")),
                ref_symbols = c("*"),
                part = "header",inline=T)
ft <- footnote( ft, i = 1, j = 1,
                value = as_paragraph(c("Results summarized here are for climate-only models with RW as response variable.")),
                ref_symbols = c("**"),
                part = "body",inline=T)
ft <- footnote( ft, i = c(10,11), j = 2,
                value = as_paragraph(c("Refers to SSC with significant (p<0.05) main effect of climate on RW.")),
                ref_symbols = c("#"),
                part = "body",inline=T)
ft <- footnote( ft, i = 19, j = 1,
                value = as_paragraph(c("Results summarized here are for models with BAI as response variable.")),
                ref_symbols = c("##"),
                part = "body",inline=T)
ft

>>>>

To realistically estimate forest woody productivity, it is necessary to measure or model the growth rate of the individual trees within the stand based on the primary biotic and environmental drivers.
Needed is an analysis framework that can capture the additive and interactive effects of climate, tree size (most commonly diameter breast height, DBH), and other environmental drivers, all of which may be best described by nonlinear functions [@muller-landau_testing_2006; @rollinson_climate_2021].
While multifactorial and sometimes non-linear individual-based analysis frameworks have been applied in tree ring analysis [e.g., @evans_fusing_2017; @klesse_amplifying_2020; @zuidema_recent_2020; @rollinson_climate_2021], their use has been relatively limited, and none simultaneously account for climate, DBH, and calendar year (a proxy for slowly changing environmental drivers).
In the following paragraphs, we outline major unresolved questions regarding the influence of these factors on tree growth that can best be address using a multifactorial, non-linear approach to tree ring analysis (Table `r hypotheses`).

Understanding the climate sensitivity of tree growth is critical to predicting forest dynamics and productivity as the climate changes. 
The classic dendrochronological approach to characterizing the climate sensitivity of tree growth describes linear relationships between the primary growth-limiting climate factor (moisture or temperature) and population-level growth responses captured in ring-width index chronologies [@fritts_tree_1976; @speer_fundamentals_2010].
While invaluable for applications such as reconstructing past climates [e.g., @buntgen_2500_2011], accurately characterizing the climate sensitivity of forest productivity requires more in-depth characterization of climate sensitivity across a representative sample of trees in a forest stand [@babst_when_2018].
For instance, the additive or interactive effects of precipitation and temperature are not commonly considered [but see e.g., @foster_predicting_2016; @meko_seascorr_2011;  @sanchez-salguero_disentangling_2015; @vlam_temperature_2014; @zuidema_recent_2020], and we hypothesize that both influence the growth of most species, often over different seasonal windows (Table `r hypotheses`).
In addition, we hypothesize that nonlinear climate responses, already known to be widespread within forest settings [@woodhouse_artificial_1999; @wilmking_global_2020; @rollinson_climate_2021], are in fact the predominant form of response in forests around the world (Table `r hypotheses`). 
Finally, while the influence of DBH is typically removed through detrending [@cook_calculating_1997], thereby eliminating the potential to directly model its influence on climate sensitivity, we hypothesize that interactive effects of DBH and climate are, in fact, quite common in forest settings [Table 1, @bennett_larger_2015; @trouillier_size_2019; @mcgregor_tree_2020; @rollinson_climate_2021].

Tree diameter, DBH, scales predictably with numerous traits affecting tree growth (e.g., height, crown size and position, root mass, hydraulic architecture) and therefore is itself linked to growth [e.g., @anderson-teixeira_sizerelated_2015].
Yet, there remain inconsistencies across disciplines (dendrochronology, forest ecology) as to what is considered a typical growth pattern. 
Tree ring records commonly show a pattern in which radial stem growth increments (ring width, RW) are initially large and decline with DBH [@fritts_tree_1976], whereas censuses of globally distributed forests indicate that diameter increments most commonly increases with DBH [@muller-landau_testing_2006; @anderson-teixeira_sizerelated_2015].
We hypothesize that this discrepancy is primarily a distinction between trees that established in the open (disproportionately sampled for tree-ring analysis), as opposed to in the understory (captured by contemporary monitoring; Table `r hypotheses`).
Building upon observed ontogenetic patterns in RW, dendrochonology studies often observe or assume that basal area increment (BAI) is constant with DBH after a juvenile growth phase [@fritts_tree_1976; @biondi_theorydriven_2008] -- a pattern that we would not expect to hold in understory-established trees (Table `r hypotheses`).
Finally, there is debate as to whether biomass increment ($\Delta$AGB) increases continuously with DBH [@stephenson_rate_2014; @meakem_role_2018; @foster_predicting_2016] or peaks and declines as trees divert carbon to other functions [e.g., reproduction, @thomas_agerelated_2011]. 
Following the finding that the latter pattern is common for individual trees whereas the former emerges in "cross-sectional" analyses of forest stands [@forrester_does_2021], we hypothesize that $\Delta$AGB -- and also BAI -- peaks and declines as DBH increases (Table `r hypotheses`).
Solid understanding of these ontogenetic growth trends is essential not only for predicting the growth rate of any given tree, but also for standardizing for DBH to deduce the influence of slowly changing environmental drivers [see next paragraph, @peters_detecting_2015], with the reliability of such analyses contingent upon accurate assumptions of ontogenetic growth patterns.

Beyond the direct effects of climate, rising atmospheric CO~2~ concentrations, changes in atmospheric deposition of sulfur dioxide (SO~2~) and nitrogen oxides (NO~x~), and indirect effects of climate change are all potentially influencing tree growth [@levesque_water_2017; @mathias_disentangling_2018; @maxwell_higher_2019; @walker_integrating_2020; @takahashi_air_2020; @belmecheri_precipitation_2021], and understanding these effects is central to predicting the future of the terrestrial carbon sink [@walker_integrating_2020].
Yet, characterizing how tree growth and forest productivity are responding to slowly changing environmental drivers is challenging and uncertain.
Ontogenetic patterns in tree growth must be accounted for, yet two of the most commonly used methods of standardizing for tree size, conservative detrending and basal area correction [@peters_detecting_2015], assume certain growth patterns unlikely to be universal in forest settings, as discussed above. 
Approaches that combine cross-sectional with temporal analyses to correct for growth ontogeny, such as regional curve standardization, perform better at growth trend detection [@peters_detecting_2015].
Yet, even after correcting for ontogenetic growth trends, growth trend detection remains subject to various potential sampling and analysis biases [@cherubini_potential_1998; @brienen_detecting_2012; @nehrbass-ahles_influence_2014; @sullivan_effect_2016; @bowman_detecting_2013; @brienen_tree_2017; @hember_tree_2019], fundamentally driven by the limitation that it is not possible to use a contemporary set of tree cores to obtain a representative sample of a species's population at all points throughout the history of a dynamically changing stand.
Stand dynamics strongly influence growth: tree growth rates are sensitive to competition, the intensity of which tends to increase as forests mature, affecting ecosystem level patterns of C allocation. 
Ecosystem-level carbon allocation to woody growth -- as opposed to leaf or fine root production, reproduction, defenses, etc. -- is known to decline as forest stands age [@goulden_patterns_2011; @pregitzer_carbon_2004; @delucia_forest_2007; @collalti_forest_2020]. 
Thus, we hypothesize that size-corrected growth rates of tree populations sampled from within secondary or severely disturbed stands (i.e., those with large recruitment pulses within the past century) will generally decline, whereas populations sampled from older, relatively undisturbed stands will display mixed growth trends that are more dependent on external environmental drivers (Table `r hypotheses`).

Here, we develop and apply a new method that allows simultaneous consideration of the effects of primary climate drivers (i.e., the most influential climate variables and the seasonal window over which they operate), tree size, and calendar year on annual tree growth.
This approach allows us to address hypotheses concerning the primary drivers of individual tree growth (Table `r hypotheses`) across ten forested sites spanning 52 degrees latitude.

@@

hypotheses_table <- subset(hypotheses_table, select = -c(Q, H, category,notes) )#remove columns that shouldn't be included
ft <- flextable(hypotheses_table)
ft <- italic(ft, i=hypothesis_rows, j=1)
ft <- bold(ft, i=c(question_rows))
ft <- line_spacing(ft, space = 1, part = "all")
ft <- padding(ft, padding.top = 2)
ft <- padding(ft, padding.bottom = 2)
ft <- padding(ft, j = 1, padding.left = 20)
ft <- padding(ft, i = hypothesis_rows, j = 1, padding.left = 10)
ft <- padding(ft, i = question_rows, j = 1, padding.left = 0)
ft <- autofit(ft)
ft <- footnote( ft, i = 1, j = 2,
                value = as_paragraph(c("SSC= species-site combinations")),
                ref_symbols = c("*"),
                part = "header",inline=T)
ft <- footnote( ft, i = 1, j = 1,
                value = as_paragraph(c("Results summarized here are for climate-only models with RW as response variable.")),
                ref_symbols = c("**"),
                part = "body",inline=T)
ft <- footnote( ft, i = c(10,11), j = 2,
                value = as_paragraph(c("Refers to SSC with significant (p<0.05) main effect of climate on RW.")),
                ref_symbols = c("#"),
                part = "body",inline=T)
ft <- footnote( ft, i = 19, j = 1,
                value = as_paragraph(c("Results summarized here are for models with BAI as response variable.")),
                ref_symbols = c("##"),
                part = "body",inline=T)
ft
